Machine Learning guided workflow\\ for Ribosome Binding Site engineering

### 2021-06-09 paper/main.tex
Updates from Overleaf

One of the main tenets of synthetic biology is design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\

@@

We present our RBS optimizing DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
in the DESIGN stage (Section~\ref{sec:ucb-results}),
and we demonstrate that the ML predictions are accurate in the LEARN stage (Section~\ref{sec:gp-results}).
We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and describe some interesting
characteristics of the discovered sequences, 
as well as show the effectiveness of the automated laboratory workflow stages.

@@

We show our DBTL workflow in Figure \ref{fig: Flowchart}, that uses machine learning to optimise protein expression.
BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
Machine learning algorithms are applied in LEARN and DESIGN.
In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.\\

@@

Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
This sequence is a known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
Since this is the sequence against which new RBS sequences will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} from now on.
We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
In effect, in our design we focus on modification of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T). 
The total experimental (variant) space is then $4^6$ = 4096.

@@

Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
In each round, we measure the TIR of benchmark RBS as the internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot_proj.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.42]{plots/Main_Paper/tsneplot.pdf}
    \end{subfigure}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{
    \textbf{TIRs of RBS groups examined in this study.} 
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate. 
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
    SD - Shine-Dalgarno sequence.
    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
    BN - benchmark sequences for all plates. 
    They are not all exactly 1 due to them being shown as separate samples rather than averages.
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group and BPS-NC was removed due to changes being made outside the core in this group. 
    \textbf{C)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration. 
    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{E)} tSNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
    The area of the circle represents the experimentally obtained TIR for measured groups.}
    \label{fig: Swarmplot and Quantplot}
\end{figure}

@@

However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results become much better, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
We see that effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with high predicted mean are coloured blue and represent exploitation, those hued red are with high predicted uncertainty and represent exploration.
We can see the RBSs with high TIRs tend to come from exploitation the design space whereas the explorative points give relatively low TIR but expand our knowledge about the unknown part of the design space.\\

@@

In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our embedding function.
We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
A number of clusters were targeted by our recommendation algorithm.
For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.

@@

Figure \ref{fig: Scatterplot} shows how our algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.
One important point to note is that the predictions are also influenced by our recommendation choices. 
In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\

Our recommendation algorithm selects new designs based on means and uncertainties calculated by our prediction algorithm.
This algorithm creates a model which takes the RBS sequences as input and predicts the mean TIR values with uncertainty level about that prediction, based on the experimental data.
For this study, we have used the Gaussian Process Regression (GPR) as our prediction method.
The explicit representation of model uncertainty provides further guide for efficient searching through large experimental space of possible sequences.\\

@@

\subsection{BUILD+TEST: Characteristics of the tested sequences}
\label{sec:characteristics-of-library}

Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in Table 1.
Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (in terms of TIR ratio).
It is generally understood that guanine rich sequences are promoting strong transcription.
This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
This result combined with the Bandits' algorithm bias towards the G rich cluster shown in figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as ones with high TIR probability.

@@

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
For one, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
Secondly, given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
Thirdly, the generalisability of the method is unknown.
We believe that the method would be useful for use in desgin of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence.
Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.\\

We have found our approach bringing machine learning and synthetic biology experts very powerful.
We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.
However, there are still some open questions that have to be answered.
For example, we are still not sure how generalisable are the used methods and if it possible to extract more biological knowledge from the algorithms.
It is also not clear how to improve the accuracy of predictions, but we believe this will occur through  a combination of bigger data sets and better algorithms.\\

@@

All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML} .
All the processed and raw data is included in the repository.
Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as separate file.

>>>>

One of the main tenets of synthetic biology is the design, evaluation and standardisation of genetic parts \cite{Brophy2014,Canton2008,Stanton2014}.
This is usually done in terms of the Design-Build-Test-Learn (DBTL) cycle, where the given genetic part or organism are continually improved by going through a number of turns of the said cycle.
This normally involves designing the DNA sequence in Computer Aided Design (CAD) software and then physically testing it in a laboratory. 
Additionally, computer modelling and prediction of part behaviour based on the designed DNA sequence or design of DNA sequence based on expected function can be used \cite{Yeoh2019,Nielsen2016}.
Most of these models are based on either the thermodynamic properties of the involved molecules (DNA, RNA, proteins, etc.) or empirically obtained values describing a relevant to a given design property, like Translation Initiation Rate (TIR) in the case of Ribosome Binding Sites (RBS) \cite{Xia1998,Chen2013,Reeve2014}.
However, de-novo design of small genetic elements is still challenging due to unknown relationships between sequence and performance of such elements.\\

@@

We present our RBS optimising DBTL workflow that uses machine learning in Section~\ref{sec:dbtl-workflow}.
Machine learning (ML) is used in two different ways: we show the efficacy of the ML recommendations
in the DESIGN stage (Section~\ref{sec:ucb-results}),
and we demonstrate that the ML predictions are accurate in the LEARN stage (Section~\ref{sec:gp-results}).
We present our new RBS sequence library in Section~\ref{sec:characteristics-of-library} and describe some interesting
characteristics of the discovered sequences, 
as well as show the effectiveness of the automated laboratory workflow stages.

@@

We show our DBTL workflow, that uses machine learning to optimise protein expression, in Figure \ref{fig: Flowchart}.
BUILD and TEST are driven chiefly by choices made by human researchers and use of automated methods.
Machine learning algorithms are applied in LEARN and DESIGN.
In LEARN phase, we use the Gaussian Process regression algorithm to predict the TIR of RBS sequences comprising the experimental space.
In DESIGN phase, we use Upper Confidence Bound multi-armed Bandit algorithm to recommend new RBS designs based on the predictions from LEARN.\\

@@

Our template RBS sequence is 20 bps long with the sequence TTTAAGA\textbf{AGGAGA}TATACA, where the highlighted nucleotides constitute the core of the RBS.
This sequence is known to have a very high TIR and comes with the pBb series plasmids \cite{Lee2011}. 
Since this is the sequence against which new RBS sequences will be benchmarked,
we will refer to this sequence as the \textit{benchmark sequence} from now on.
We have experimentally confirmed that modifying the core sequence is statistically more impactful on TIR than changes made outside of it (see Supplementary materials).
In effect, in our design we focus on modification of the core at positions -8 to -13 (relative to the start codon of the GFP) nucleotides of the RBS and fix others to be the same as the benchmark sequence, i.e. TTTAAGA + NNNNNN + TATACAT, where N can be any nucleotide (A, C, G, T). 
The total experimental (variant) space is then $4^6$ = 4096.

@@

\mengyan{Maciej: do we want to put the group description into Fig 2 as we discussed last Fri?}
\maciej{I think we have decided to put it in the text, but within the description of Fig 2A as it is now. Otherwise the caption would be suuuppppppeeeerrrr long}
Figure \ref{fig: Swarmplot and Quantplot}A shows the results for all the examined groups. 
In each experimental round, we measure the TIR of benchmark RBS as the internal standard. 
We then obtain the normalised TIR (called \textit{TIR ratio}) by taking the ratio between the raw TIR and the average TIR of benchmark sequences in each round (which are run in triplicate in each round).\\

\begin{figure}[!ht]
    \centering
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/quantplot.pdf}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[b]{0.25\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.35]{plots/Main_Paper/swarmplot_proj.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.4]{plots/Main_Paper/histogram.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.48\textwidth}
        \centering
        \caption{}
        \includegraphics[scale=0.42]{plots/Main_Paper/tsneplot.pdf}
    \end{subfigure}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/quantplot.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/swarmplot_proj.pdf}
    % \includegraphics[scale=0.3]{plots/Main_Paper/histogram.pdf}
    % \includegraphics[scale=0.4]{plots/Main_Paper/tsneplot.pdf}
    \caption{
    \textbf{TIRs of RBS groups examined in this study.} 
    \textbf{A)} Swarm plot showing the obtained TIRs divided into RBS groups.
    BPS-NC: base-by-base changes in the non-core region. 
    BPS-C: base-by-base changes in the core region. 
    UNI: Randomly generated sequences with uniform distribution. 
    PPM: Randomly generated sequences with distribution following the PPM for all natural RBS in \emph{E. coli}. 
    Bandit-0/1/2/3 - Bandit algorithm generated results for Round 0, 1, 2 and 3 respectively.
    SD - Shine-Dalgarno sequence.
    Dash line is set to 1 and represents the averaged benchmark sequence TIR for that group. 
    BN - benchmark sequences for all plates. 
    They are not all exactly 1 due to them being shown as separate samples rather than per round averages.
    \textbf{B)} Line plot showing TIR obtained in a given quantile of results divided into groups as in A).
    % save the random groups which were shown together due to similar distributions.
    UNI and PPM are merged into Random group and BPS-NC is not shown due to changes being made outside the core in that group. 
    \textbf{C)} Exploitation v.s. Exploration for Bandit 1-3. Blue-hued points represent exploitation, those hued red represent exploration. 
    \textbf{D)} Histogram with kernel density estimations (KDE) showing distributions of TIRs for Bandit groups.
    \textbf{E)} t-SNE plot showing the relative distances between sequences in our design spaces as calculated by our kernel function (weighted degree kernel with shift). 
    The area of the circle corresponds to the the experimentally obtained TIR value.
    The TIR results in all subplots are shown normalised to the respective benchmark sequence sample which acts as internal standard, that is TIR of a given RBS is divided by TIR of the benchmark RBS run in the same plate. }
    \label{fig: Swarmplot and Quantplot}
\end{figure}

@@

However, starting from Round 1, where the bandit algorithm was fed data from the Round 0 the results improve significantly, with a number of sequences that perform better than the consensus Shine-Dalgarno sequence and in one case better than the benchmark (by 8\%).
In round 2 we have observed further improvement by getting more sequences that showed TIR on levels similar to our benchmark sequence.
Finally, in round 3 the algorithm identified two sequences that were 34\% and 15\% stronger than the benchmark sequence.\\

Figure \ref{fig: Swarmplot and Quantplot}B shows the same results but divided into quantiles where the specific point for a given group is showing the highest TIR for that quantile.
The gradual increase for all quantiles can be observed for all Bandit groups suggesting algorithms' better understanding of the experimental space with more data.
The decreased result in the 0.9th qunatile compared to the max value for Bandit 3 group can be attributed to the increased emphasis on exploitation that has been set for the that round compared to others.
We see that effect in Figure \ref{fig: Swarmplot and Quantplot}C, where we coloured the data points for Bandit 1-3 groups according to their relative exploration - exploitation affinity.
Those with high predicted mean are coloured blue and represent exploitation, those hued red are with high predicted uncertainty and represent exploration.
We can see that the RBSs with high TIRs tend to come from exploitation of the design space whereas the explorative points give relatively low TIR, but expand our knowledge about the unknown part of the design space.\\

@@

In \ref{fig: Swarmplot and Quantplot}E we show a t-SNE plot depicting the experimental space.
Each RBS is located on the plot according to its distance to other RBSs as calculated by our embedding function.
We can see the RBSs recommended by Bandit groups have covered majority of the design space. 
Additionally, a number of clusters were especially targeted by our recommendation algorithm.
For example, the circled clusters labelled as ``G-Rich Clusters" have been actively recommended by the algorithm.
More specifically, sequences with more or equal to 4 guanines in any position constituted 10\% of the randomly selected sequences and 5, 9, 16 and finally 25\% in each of the 4 Bandit guided batches respectively.

@@

Figure \ref{fig: Scatterplot} shows how our implemetation of the Gaussian Process algorithm performed in terms of predictions in each round. 
As expected, the predictions in Round 0 were poor due to use of approximated data. 
The predictions improved for the subsequent rounds, from R\textsuperscript{2} of 0.065 for round 0 to R\textsuperscript{2} of 0.27 for round 3.
Similarly, the Spearman correlation coefficient rose from 0.27 for Round 0 to 0.48 for Round 3.\\

One important point to note is that the predictions are also influenced by our recommendation choices. 
In each round, we select a number of data points for exploration, which means that these data points, when tested, have a high chance of having real mean different to what was predicted.
However, this is still very useful information for future predictions as it allows us to understand the underlying space better.\\

@@

\subsection{BUILD \& TEST: Characteristics of the tested sequences}
\label{sec:characteristics-of-library}

Our data set taken together can be viewed as a reliable library of RBS sequences for \emph{E. coli}, some characteristics of which are shown in Table 1.\\

Figure \ref{fig:Library characteristics}A shows the sequence logo calculated for the Top 30 sequences (in terms of TIR ratio).
It is generally understood that guanine rich sequences are promoting strong transcription.
This expected bias towards guanine is clearly visible for all positions in our Top 30 RBSs.
This result combined with the Bandits' algorithm bias towards the G rich cluster shown in Figure \ref{fig: Swarmplot and Quantplot}D reinforces the notion that our algorithm successfully identified G rich sequences as the ones with high TIR probability.\\

@@

We have found our approach bringing machine learning and synthetic biology experts very powerful.
We envision that pairing of machine learning with high-throughput automation will keep delivering a high number of good quality data sets and improved methods for biological engineering.\\

There are still open questions that need to be addressed for applying machine learning in synthetic biology.
For one, we would like to understand how can we extract more biologically important information from the decisions made by our algorithms.
Secondly, given a small amount of RBS sequences tested, how can machine learning algorithm provide more accurate predictions and uncertainty measurement? 
Thirdly, the generalisability of the method is unknown.
We believe that the method would be useful for use in design of other small genetic parts, but the complexity of the task quickly increases with the size of the analysed sequence and so the method's applicability might be impacted at some point.
Finally, the optimal balance between exploration-exploitation between rounds and samples is still to be determined.
In this research we have decided to gradually move the balance from exploration to exploitation over the rounds, but it is not yet clear what is the method protocol for this change.\\

@@

All code and data required to reproduce the results is available at Github: \url{https://github.com/mholowko/SynbioML} .
All the processed and raw data is included in the repository.
Sequences of plasmids and oligos and assembly reports used in this study are available in supplementary as a separate file.
